---
title: "Krill density time series"
author: "Cox, M.J., Reid, K."
date: "22/06/2020"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This is a preliminary analysis of the krill density time series made up of data submitted to CCAMLR WG-ASAM-20.

## Set-up
Load ```R``` packages:
```{r loadpkg}
suppressPackageStartupMessages({library(readxl)
library(tibble)
library(dplyr)
library(zoo)
library(ggplot2)
  library(cowplot)})

```

Load data and user defined variables:
To load the time series data from excel put all the data files in one directory.  The directory is specified in the following code chunk ```data.files$wd```.

```{r usrVar}

data.files=list(
meta=tibble(locName=c('SG',
                      'SO',
                      'WAP',
                      'SSH',
                      'WAP2'),
            fn=c('BAS_Antarctic_krill_density_South_Georgia_20200615.xlsx',
            'IMR_Antarctic_krill_density_SouthOrkneys_2020.xlsx',
            'WG_EMM_krill_biomass_time_series_information_GW.xlsx',
            'WG_EMM_krill_biomass_time_series_information_China.xlsx',
            'WG_EMM_krill_biomass_time_series_information_PS112-2-3.xlsx')))


```

In the above code chunk, the object ```data.files``` is a ``list`` with two elements: ```$wd``` is the directory where the excel krill density files are found (this could be changed to a server location) and ```$meta``` gives a simple file designator name ```$locName``` and ```$fn``` ia the Excel file name.

Now we can load the data into the ```R``` workspace:

```{r loadData,message=FALSE,warning=FALSE}
dat=lapply(as.list(paste0(data.files$wd,data.files$meta$fn)),function(x) readxl::read_xlsx(x,skip=2))
names(dat)=data.files$meta$locName

```
Each file is now in an element of the ```dat``` ```list``` object.

Now we'll look at the number of rows and columns in each file:

```{r size}
dat=lapply(as.list(paste0(data.files$wd,data.files$meta$fn)),function(x) readxl::read_xlsx(x,skip=2))
names(dat)=data.files$meta$locName

```

```{r sizeTable,echo=FALSE,message=FALSE}
dispSize=tibble(locName=data.files$meta$locName,nrow=sapply(dat,nrow),
             ncol=sapply(dat,ncol))

knitr::kable(dispSize,caption = 'Size of each of the data files')

```


From Table \@ref(tab:sizeTable) we see that the 'SG' file has two extra columns, so we remove remove those to enable the data to be merged together:

```{r rmCols}
#remove extra columns in SG
dat[[1]]=dat[[1]][,-c(23,24)]
```

Now we merge the data together into a single ```R``` object:

```{r mergeDat}
### change SO to numeric
dat$SO$`Density estimate (g m-2)` <- as.numeric(dat$SO$`Density estimate (g m-2)`)
dat$SO$`Survey area (km2)` <- as.numeric(dat$SO$`Survey area (km2)`)


#dat=do.call('rbind.data.frame',dat)
dat =  dplyr::bind_rows(dat)
nall=nrow(dat) #number of surveys

```
The entire data set has ```r nall``` rows.

## Filter data
In the following sets we remove some rows from the data and provide the reason for removing the data:

```{r filtData}
filterStep=c('None, this is all the data','remove blank densities')
## remove blank densities

dat=dat %>% filter(!is.na(`Density estimate (g m-2)`))
nV=c(nall,nrow(dat))

filterStep=c(filterStep,'Remove blank CV')
dat=dat %>% filter(!is.na(`CV of density estimate (%)`))
nV=c(nV,nrow(dat))

## remove unspecified variance estimate
filterStep=c(filterStep,'Remove blank variance est. meth.')
dat=dat %>% filter(!is.na(`CV estimation method`))
nV=c(nV,nrow(dat))
```

The data filtering is summarised in Table \@ref(tab:filtTab).

```{r filtTab,echo=FALSE}
nTab=tibble(filter=filterStep,nRm=nV)
names(nTab)=c('Filter step','Number of observations remaining')
knitr::kable(nTab,caption='Data filtering steps and remaining number of samples.')
```

## Use a single contributor per data set
Here we use a single contributor per data set (row in ```dat```):

```{r singleContrib}
ind=grep(',',dat$Contributor)
dat$Contributor[ind]='IMR'

dat$Contributor[dat$Contributor=='IMR']='Norway'
```

## Create a time stamp
Here we create a time stamp to help with plotting:

```{r timeStamp}
dat$timeStamp=zoo::as.yearmon( paste(dat$`Month (MON)`,dat$`Year (yyyy)`),format='%b %Y')
```

## Prepare data for plotting
```{r prepData}
#make numeric density estimates:
dat$`Density estimate (g m-2)`=as.numeric(dat$`Density estimate (g m-2)`)
#make factor variable for subarea:
dat$SubareaF=paste0('Subarea',dat$Subarea)

dat$SubareaF=factor(paste0('Subarea',dat$Subarea),levels=paste0('Subarea',c(48.1,48.2,48.3)))

# standard deviation of rho   
dat$SD=dat$`Density estimate (g m-2)` * 0.01 * dat$`CV of density estimate (%)`
# variance of rho
  dat$Var=dat$SD**2
# 95% confidence assuming normal distribution:  
  dat$LB=qnorm(p=0.025,mean=dat$`Density estimate (g m-2)`,sd=dat$SD)
  dat$UB=qnorm(p=0.975,mean=dat$`Density estimate (g m-2)`,sd=dat$SD)
               
#prep month:
  dat$mon=dat$`Month (MON)`
dat$monF=factor(dat$mon,levels=month.abb)
  
```

# Results
(Table \@ref(tab:subAreaByNation))

```{r subAreaByNation,echo=FALSE}
knitr::kable(table(dat$Subarea,dat$Contributor), caption='Area by contributor')
```

## Time series plots

Number of surveys for each year by subarea (Fig. \@ref(fig:sampleSize))

```{r sampleSize,fig.cap='Number of surveys carried outduring each year by subarea.',echo=FALSE}

sampSize= dat %>% group_by(SubareaF,`Year (yyyy)`) %>%
  summarise(n=n())

psize=ggplot(data=sampSize,aes(x=`Year (yyyy)`,
                               y=n,
                               group=SubareaF)) +
  geom_point() + facet_wrap(~SubareaF,ncol=1) +
  theme_minimal() + xlab('Year') + 
  ylab('Number of surveys')


psize=ggdraw(psize) + 
  draw_label("Draft", color = "grey", size = 110, angle = 45,alpha=0.5)
print(psize)
```

```{r basePlot,echo=FALSE}
p1=ggplot(dat,aes(x=timeStamp,
                  y=`Density estimate (g m-2)`,
                  group=SubareaF,colour=SubareaF,
                  shape=Contributor)) + 
  geom_point(alpha=0.8) + 
 xlab('Year') +
  theme_minimal() + 
  facet_wrap(~SubareaF,ncol=1) +
  ylab(bquote('Areal krill biomass density, '~gm^-2)) + 
   labs(colours='Subarea') 
```

```{r normCI,echo=FALSE,fig.cap='Krill density time series by subarea.  The shaded area is the 95% confidence interval assuming normally distributed errors.  The lower bound of the 95% confidence interval is sometimes less than zero which is not ideal for representing densitty. NB y-axis limits are set  -50 to 300 gm-2'}
pNCI=p1+geom_errorbar(aes(x=timeStamp,ymin=LB,ymax=UB,colour=SubareaF,group=SubareaF,fill=SubareaF),alpha=0.8) +
  #geom_ribbon(aes(x=timeStamp,ymin=LB,ymax=UB,colour=SubareaF,group=SubareaF,fill=SubareaF),alpha=0.2)  +
  geom_hline(yintercept=0,colour='grey') +  coord_cartesian(ylim=c(-50,300)) + NULL #ylim(c(0,260)) 

pNCI=ggdraw(pNCI) + 
  draw_label("Draft", color = "grey", size = 110, angle = 45,alpha=0.5)
print(pNCI)

```


We now represent the distribution of densities for each survey using log-normal confidence intervals.  We use ```R``` code *Borchers, D. L., & Cox, M. J. (2017). Distance sampling detection functions: 2D or not 2D?. Biometrics, 73(2), 593-602.*:

```{r logNorm}
lnci.nmin=function(stat,cv,stat.min=0,interval=interval){
  q=Mod(qnorm((1-interval)/2,0,1))
  varNhat=(stat*cv)^2
  cfactor=exp(q*sqrt(log(1+varNhat/(stat-stat.min)^2)))
  lower=stat.min+(stat-stat.min)/cfactor
  upper=stat.min+(stat-stat.min)*cfactor
  return(list(lower=lower,upper=upper))
}
```

and run ```lnci.nmin()```:

```{r logNormRun}

logCI=lnci.nmin(stat=dat$`Density estimate (g m-2)`,
          cv=dat$`CV of density estimate (%)`/100,
          interval=0.95)
dat$LBlnn=logCI$lower
dat$UBlnn=logCI$upper

```


```{r logNplot,echo=FALSE,fig.cap='Krill density time series by subarea.  The shaded areas are the 95% confidence interval assuming a log-normal error distribution.  NB y-axis limits are set  [-50 to 300 gm-2]'}
p4=p1+geom_errorbar(data=dat,
                 aes(x=timeStamp,ymin=LBlnn,ymax=UBlnn,
                   colour=SubareaF,group=SubareaF),
                   alpha=0.8) +
  #geom_ribbon(data=dat,aes(x=timeStamp,ymin=LBlnn,ymax=UBlnn,
                   #colour=SubareaF,group=SubareaF,
                   #fill=SubareaF),alpha=0.2) +
  coord_cartesian(ylim=c(-50, 300)) + 
  geom_hline(yintercept=0,col='grey')

p5=ggdraw(p4) + 
  draw_label("Draft", color = "grey", size = 110, angle = 45,alpha=0.5)
print(p5)

ggsave('SubareaLogNormal95pcConfidence.png')

```
## Changes in CV with month and equipment
Krill behaviour, e.g. swarming and environmental preference, will influence CV,i.e. evenly spaced krill will reduce CV, whereas a single massive 'super' swarm will greatly increase CV.  It is reasonable to see if CV changes with observation month, so we take a look at the in Fig. \@ref(fig:CVmonth).

```{r CVmonth,echo=FALSE,fig.cap='Coefficient of variation by subarea and observation month. Sample size for each boxplot is given.'}

p6=ggplot(dat,aes(x=monF,
                  y=`CV of density estimate (%)`,
                  group=monF)) + 
  geom_boxplot() + facet_wrap(~SubareaF,ncol=1) +
  stat_summary(fun = median, fun.max = length,
                geom = "text", aes(label = ..ymax..), vjust = -1) + theme_minimal() + xlab('Month')

p6=ggdraw(p6) + 
  draw_label("Draft", color = "grey", size = 110, angle = 45,alpha=0.5)
print(p6)

ggsave('SubareaCVByMonth.png')

```

And now by Echosounder:
```{r boxES,echo=FALSE,fig.cap='Coefficient of variation by subarea and Echosounder type.  Sample size for each boxplot is given.'}

p7=ggplot(dat,aes(x=`Echosounder model`,
                  y=`CV of density estimate (%)`,
                  group=`Echosounder model`)) + 
  geom_boxplot() + facet_wrap(~SubareaF,ncol=1) +
  stat_summary(fun = median, fun.max = length,
                geom = "text", aes(label = ..ymax..), vjust = -1) + theme_minimal() + xlab('Month')

p7=ggdraw(p7) + 
  draw_label("Draft", color = "grey", size = 110, angle = 45,alpha=0.5)
print(p7)

ggsave('SubareaCVByMonth.png')
```


## Unweighted average and variance
```{r unwtAvg}
summaryStatTime=dat %>% group_by(SubareaF) %>% summarise(n=n(),
  start=min(`Year (yyyy)`),
  end=max(`Year (yyyy)`),
  mean=mean(`Density estimate (g m-2)`),
  var=sqrt(sum(Var)),
  SD=sqrt(var),
  CV=SD/mean)

```


```{r summaryStatTime,echo=FALSE}
knitr::kable(summaryStatTime,caption='Summary statistics for each time series using unweighted observations.')
```